---
title: "10 Deep Learning"
author: "By: Udit (based on ISLR)"
geometry: "left=1cm,right=1cm,top=1cm,bottom=2cm"
output: pdf_document
---

## Fitting Linear & Lasso Models

The **pipe operator %>%** passes the previous term keras model sequential pipe as 
the first argument to the next function, and returns the result.

```{r}
library(ISLR2)
library(ggplot2)
library(magrittr) # for pipe operator
library(keras)

names(Hitters)
summary(Hitters)
hit.data = na.omit(Hitters)

# Split data into test and train
n = nrow(hit.data)
set.seed(13)
test = sample(1:n, n/3)

# Fitting linear model
ln.fit = lm(Salary~., data=hit.data[-test,])
ln.pred = predict(ln.fit, hit.data[test,])
sqrt(mean((ln.pred - hit.data$Salary[test])^2))  # RMSE = 341

# Fitting lasso using glmnet - need to create model matrix
library(glmnet)
x = model.matrix(Salary~. -1 , data=hit.data) %>% scale()
y = hit.data$Salary

cvfit = cv.glmnet(x[-test,], y[-test], type.measure = "mae")
cvpred = predict(cvfit, x[test,], s = "lambda.min")
sqrt(mean((cvpred - hit.data$Salary[test])^2))  # RMSE = 359
```

## Fitting Neural Network

The object modnn has a single hidden layer with 50 hidden units, and a ReLU activation 
function. It then has a dropout layer, in which a random 40% of the 50 activations 
from the previous layer are set to zero during each iteration of the stochastic gradient 
descent algorithm. Finally, the output layer has just one unit with no activation 
function, indicating that the model provides a single quantitative output.  

**units** - dimensionality of output space.  
**input_shape** - Dimensionality of the input (integer) not including the 
samples axis. This argument is required when using this layer as the first layer in a model.

```{r}
# Creating a network and adding details - o/p is single quantitative output
modnn = keras_model_sequential() %>% 
  layer_dense(units=50, activation="relu", input_shape = ncol(x)) %>%
  layer_dropout(rate = 0.4) %>%
  layer_dense(units = 1)
summary(modnn)

# Add details on fitting algorithm (compile passes the info to python instance)
modnn %>% compile(loss="mse", optimizer = optimizer_rmsprop(), 
                  metrics = list("mean_absolute_error"))

# Fit the model - 2 parameters (epochs and batch_size)
history = modnn %>% fit(x[-test,], y[-test], epochs=1500, batch_size=32,
                        validation_data = list(x[test,], y[test]))
plot(history)
npred = predict(modnn, x[test,])
sqrt(mean((npred - hit.data$Salary[test])^2))  # RMSE = 339
```

## Fitting Neural Network - MNIST Digit Data

There are 60,000 images in the training data and 10,000 in the test data.
The images are 28x28, and stored as a 3D array.

Neural networks are somewhat sensitive to the scale of the inputs. Here the inputs
are eight-bit grayscale values between 0 and 255, so we scale to the unit interval.
```{r}
mnist = dataset_mnist()
x_train = mnist$train$x
y_train = mnist$train$y
dim(x_train)
x_test = mnist$test$x
y_test = mnist$test$y
dim(x_test)

get_matrix = function(x){
  array_reshape(x, c(nrow(x), 28*28))
}

x_train = get_matrix(x_train)
x_test  = get_matrix(x_test)

# example
to_categorical(head(y_train), 10)

y_train = to_categorical(y_train, 10) #convert to categorical w/ 10 classes
y_test  = to_categorical(y_test, 10)


# scaling - Neural networks are sensitive to scale
x_train = x_train/255
x_test  = x_test/255

# Create a Neural Network model
modelnn <- keras_model_sequential() 
modelnn %>% 
  layer_dense(units = 256, activation = 'relu', input_shape = c(784)) %>% 
  layer_dropout(rate = 0.4) %>% 
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dropout(rate = 0.3) %>%
  layer_dense(units = 10, activation = 'softmax')
summary(modelnn)

# Details for fitting
modelnn %>% compile(
  loss = 'categorical_crossentropy',
  optimizer = optimizer_rmsprop(),
  metrics = c('accuracy')
)

# Fit the model
system.time(
history <- modelnn %>% fit(x_train, y_train, epochs=30, batch_size=128,
                            validation_split=0.2))
plot(history, smooth=FALSE)

# Calculate accuracy
accu = function(pred, truth) {mean(drop(pred)== drop(truth))}

ypred = modelnn %>% predict(x_test) %>% k_argmax()
accu(ypred$numpy(), mnist$test$y)  #98%

# Fitting a single layer model
modellr = keras_model_sequential() %>% layer_dense(input_shape=784, units=10, 
                                                   activation="softmax")
summary(modellr)

modellr %>% compile(loss="categorical_crossentropy", 
                     optimizer=optimizer_rmsprop(),
                     metrics=c("accuracy"))
modellr %>% fit(x_train, y_train, epochs=30, batch_size=128, validation_split=0.2)
ypred = modellr %>% predict(x_test) %>% k_argmax()
accu(ypred$numpy(), mnist$test$y)  #90%
``` 

## CNN - Convolutional Neural Network

The array of 50,000 training images has 4 dimensions: each color image is represented
as a set of 3 channels, each of which consists of 32x32 8bit pixels.

```{r}
#cifar = dataset_cifar100()
setwd("C:/Users/uditg/Documents/R scripts")
cifar = readRDS("cifar100_object")
names(cifar)

x_train <- cifar$train$x
g_train <- cifar$train$y
x_test  <- cifar$test$x
g_test  <- cifar$test$y
dim(x_train)
range(x_train[1,,,1])

x_train = x_train/255
x_test  = x_test/255
y_train = to_categorical(g_train,100)
dim(y_train)

# plotting sample images
library(jpeg)
par(mar=c(0,0,0,0), mfrow=c(5,5))
index = sample(seq(50000),25)
for (i in index) plot(as.raster(x_train[i,,,]))

# Moderately sized CNN model
modelcnn = keras_model_sequential() %>% 
  layer_conv_2d(filters=32, kernel_size=c(3,3),
                padding = "same", activation="relu",
                input_shape = c(32, 32, 3)) %>%
  layer_max_pooling_2d(pool_size = c(2,2)) %>%
  layer_conv_2d(filters=64, kernel_size=c(3,3),
                padding="same", activation="relu") %>%
  layer_max_pooling_2d(pool_size = c(2,2)) %>%
  layer_conv_2d(filters=128, kernel_size=c(3,3),
                padding="same", activation="relu") %>%
  layer_max_pooling_2d(pool_size = c(2,2)) %>%
  layer_conv_2d(filters=256, kernel_size=c(3,3),
                padding="same", activation="relu") %>%
  layer_max_pooling_2d(pool_size = c(2,2)) %>%
  layer_flatten() %>%
  layer_dropout(rate=0.5) %>%
  layer_dense(units=512, activation="relu") %>%
  layer_dense(units=100, activation="softmax")
summary(modelcnn)

modelcnn %>% compile(loss="categorical_crossentropy", optimizer=optimizer_rmsprop(),
                  metrics=c("accuracy"))
history = modelcnn %>% fit(x_train, y_train, epochs=30,
                        batch_size=128, validation_split=0.2)
ypred = modelcnn%>% predict(x_test) %>% k_argmax()
accu(ypred$numpy(), g_test)  #45%
```

## CNN Pretrained Models - ImageNet

```{r}
img_location = "CNN_images"
image_names = list.files(img_location)
num_images = length(image_names)
x = array(dim=c(num_images, 224, 224, 3))
for(i in 1:num_images){
  img_path = paste(img_location, image_names[i], sep="/")
  img = image_load(img_path, target_size=c(224,224))
  x[i,,,] = image_to_array(img)
}
x = imagenet_preprocess_input(x)

modelpre = application_resnet50(weights="imagenet")
#summary(modelpre)
pred6 = modelpre %>% predict(x) %>% imagenet_decode_predictions(top=3)
names(pred6) = image_names
print(pred6)
```

## IMDb Document Classification

Using:
1. Logistic regression w/ Lasso regularization
2. Bag-of-words model
3. **RNN (handles vector embedding too)**

#### Logistic regression w/ Lasso regularization
We score each document for the presence or absence of each of the words in
a language dictionary - in this case an English dictionary. If the dictionary
contains M words, that means for each document we create a binary feature
vector of length M, and score a 1 for every word present, and 0 otherwise.

```{r}
max_features = 10000
imdb = dataset_imdb(num_words=max_features)
max(unlist(x_train))
c(c(x_train, y_train), c(x_test, y_test)) %<-% imdb

x_train[[1]][1:12]

word_index <- dataset_imdb_word_index()

# Text codes are off by 3 because of adjustments, made explicitly below
word = names(word_index)  #words in the dictionary
idx  = unlist(word_index, use.names=FALSE) #paired values
word = c("<PAD>","<START>","<UNK>","<UNUSED>",word) #appending values
idx  = c(0:3, idx+3)
  
decode_review = function(text){
  words= word[match(text, idx, 2)] #returns 2 when no match
  paste(words, collapse = " ")
}

decode_review(x_train[[1]][1:12])
y_train[1]  # 1 = Good
decode_review(x_train[[3]][1:12])
y_train[3]  # 0 = Bad

# One-hot encoding
library(Matrix)
one_hot <- function(sequences, dimension){
  # create 'i' - row #
  seqlen = sapply(sequences, length)
  n = length(seqlen)
  row.ind = rep(1:n, seqlen)
  
  # create 'j' - column #
  col.ind = unlist(sequences)
  
  #i,j specify location of non-zero elements
  sparseMatrix(i=row.ind, j=col.ind, dims=c(n, dimension)) 
}


x_train_1h = one_hot(x_train, 10000)
x_test_1h = one_hot(x_test, 10000)
dim(x_train_1h)
nnzero(x_train_1h)/(25000*10000) #only 1.3% contains non-zero value (i.e. 1)
```

First we fit a lasso logistic regression model using glmnet() on the training
data, and evaluate its performance on the validation data. Finally, we plot
the accuracy, acclmv, as a function of the shrinkage parameter, $\lambda$.

```{r}
library(glmnet)

#Validation set of 2000... 23000 for training
set.seed(3)
ival = sample(seq(along=y_train), 2000)

fitlm = glmnet(x_train_1h[-ival,], y_train[-ival], family = "binomial", 
               standardize = FALSE) #alpha =1 by default, Lasso
classlm = predict(fitlm, x_train_1h[ival,]) > 0 
acclm = apply(classlm, 2, accu, y_train[ival]>0)  #y_train is argument to accu

par(mar=c(4,4,4,4), mfrow=c(1,2))
plot(-log(fitlm$lambda), acclm)

# performance on test data: ~85%
classlm.test = predict(fitlm, x_test_1h) > 0 
acclm.test = apply(classlm.test, 2, accu, y_test>0)
plot(-log(fitlm$lambda), acclm.test)
```
#### Bag-of-words model

Next we fit a fully-connected neural network with two hidden layers, each
with 16 units and ReLU activation

```{r}
modelbow = keras_model_sequential() %>%
  layer_dense(units=16, activation="relu", input_shape=c(10000)) %>%
  layer_dense(units=16, activation="relu") %>%
  layer_dense(units=1, activation="sigmoid")

modelbow %>% compile(optimizer="rmsprop", loss="binary_crossentropy", 
                     metrics=c("accuracy"))
history = modelbow %>% fit(x_train_1h[-ival,], y_train[-ival],
                           epochs=20, batch_size=512,
                           validation_data=list(x_train_1h[ival,],y_train[ival]))
plot(history)
```

#### RNN - Recurrent Neural Network

Sentiment analysis with IMDb data.

```{r}
wc = sapply(x_train, length)
median(wc)      #median word count 178
mean(wc<=500)   #92% of reviews have <500 words

# keeps last 500 words/ adds padding in the beginning
x_train <- pad_sequences(x_train, maxlen=500)
x_test  <- pad_sequences(x_test , maxlen=500)
dim(x_train); dim(x_test)
```

At this stage, each of the 500 words in the document is represented using an integer
corresponding to the location of that word in the 10,000-word dictionary.
The **first layer** of the RNN is an embedding layer of size 32, which will be
learned during training. This layer **one-hot encodes each document as a
matrix of dimension 500×10,000**, and then maps these 10,000 dimensions
down to 32.

```{r}
modelrnn = keras_model_sequential() %>%
  layer_embedding(input_dim = 10000, output_dim = 32) %>%
  layer_lstm(units=32) %>%
  layer_dense(units=1, activation="sigmoid")

modelrnn %>% compile(optimizer="rmsprop", loss="binary_crossentropy",
                     metrics=c("acc"))
history = modelrnn %>% fit(x_train, y_train, epochs=10,
                           batch_size=128,
                           validation_data=list(x_test, y_test))
plot(history)
predy = predict(modelrnn, x_test) > 0.5
mean(abs(y_test == as.numeric(predy)))

```


## Time Series Prediction

1. AR model
2. **RNN**


#### AR model

```{r}
library(dplyr)

xdata=data.matrix(NYSE[,c("DJ_return", "log_volume", "log_volatility")])
istrain = NYSE[,"train"]  # True/ False
xdata = scale(xdata)

volume = data.frame(logvol = xdata[,"log_volume"],
                    L1 = lag(xdata,1), L2 = lag(xdata,2),
                    L3 = lag(xdata,3), L4 = lag(xdata,4), L5 = lag(xdata,5))
colnames(volume) = c("logvol", "L1.ret", "L1.volm", "L1.vol", "L2.ret", "L2.volm",
                     "L2.vol", "L3.ret", "L3.volm", "L3.vol", "L4.ret", "L4.volm",
                     "L4.vol", "L5.ret", "L5.volm", "L5.vol") 

volume = volume[-(1:5),]
istrain = istrain[-(1:5)]

#Fitting Linear AR model
arfit = lm(logvol ~., data=volume[istrain,])
arpred= predict(arfit, volume[!istrain,])
summary(arfit)$r.squared

# R-squared on test data ~41%
1 - mean((arpred - volume[!istrain, "logvol"])^2)*(1/var(volume[!istrain, "logvol"]))

volume.wk = data.frame(day=NYSE[-(1:5),"day_of_week"], volume)
arfit.wk = lm(logvol~., data=volume.wk[istrain,])
arpred.wk = predict(arfit.wk, volume.wk[!istrain,])

# R-squared on test data ~46%
1 - mean((arpred.wk - volume[!istrain, "logvol"])^2)*(1/var(volume[!istrain, "logvol"]))
```

#### RNN - Recurrent Neural Network

Need to reshape the data, for RNN since it expects a sequence of L = 5 feature 
vectors for each observation.

Two forms of dropout for the units feeding into the hidden layer. The first is 
for the input sequence feeding into this layer, and the second is for the previous
hidden units feeding into the layer. The output layer has a single unit for the response.

```{r}
n = nrow(volume)
xrnn = data.matrix(volume[,-1])
xrnn = array(xrnn, c(n, 3, 5))
dim(xrnn)

xrnn = xrnn[,,5:1]  # reversing order of lag data... index 1 is back
xrnn = aperm(xrnn, c(1,3,2)) # Transpose an array by permuting its dimensions
dim(xrnn)

modelrnn = keras_model_sequential() %>%
  layer_simple_rnn(units=12, input_shape = list(5,3), dropout=0.1,
                   recurrent_dropout = 0.1) %>%
  layer_dense(units=1)

modelrnn %>% compile(optimizer=optimizer_rmsprop(), loss="mse")

history = modelrnn %>% fit(xrnn[istrain,,], volume[istrain,"logvol"],
                           batch_size = 64, epochs = 200,
            validation_data=list(xrnn[!istrain,,], volume[!istrain, "logvol"]))

kpred = predict(modelrnn, xrnn[!istrain,,])

#40% 
1 - mean((kpred-volume[!istrain, "logvol"])^2)/var(volume[!istrain, "logvol"])


# Fitting linear AR using RNN framework
modelt = keras_model_sequential() %>% layer_flatten(input_shape=c(5,3)) %>%
  layer_dense(units=1)
modelt %>% compile(optimizer=optimizer_rmsprop(), loss="mse")

history = modelt %>% fit(xrnn[istrain,,], volume[istrain,"logvol"],
                           batch_size = 64, epochs = 50,
            validation_data=list(xrnn[!istrain,,], volume[!istrain, "logvol"]))
kpred = predict(modelt, xrnn[!istrain,,])

#41%
1 - mean((kpred-volume[!istrain, "logvol"])^2)/var(volume[!istrain, "logvol"])

```



