---
title: "11 Survival Analysis"
author: "By: Udit (based on ISLR)"
geometry: "left=1cm,right=1cm,top=1cm,bottom=2cm"
output: pdf_document
---

## Setup

Using **Survival** library.  
Using **BrainCancer** dataset.  
Function **Surv()** for creating a survival object, and **survfit() (both fitting 
& predicting) and coxph()** for model fitting.  

```{r}
library(survival)
library(ISLR2)
names(BrainCancer)
dim(BrainCancer)
attach(BrainCancer)
table(sex)
table(status)
```

## Kaplan-Meier Survival Curves

Status = 1 indicates an uncensored observation, and status = 0 indicates
a censored observation.

```{r}
fit.surv = survfit(Surv(time, status)~1)
plot(fit.surv, xlab="Months", ylab="Est. Probability of Survival")
summary(fit.surv)

fit.sex = survfit(Surv(time, status)~sex)
plot(fit.sex, xlab="Months", ylab="Est. Probability of Survival", col=c(2,4))
legend("bottomleft", levels(sex), col=c(2,4), lty=1)

# Log-Rank test
logrank.test = survdiff(Surv(time, status)~sex)
logrank.test  # p-value of 0.2 indicates null cannot be rejected
```

## Cox Proportional Hazard model

The diagnosis variable has been coded so that the baseline corresponds to meningioma. 
The results indicate that the risk associated with HG glioma is more than eight times 
(i.e. $e^{2.15}= 8.62$) the risk associated with meningioma. In addition, larger values
of the Karnofsky index, ki, are associated with lower risk, i.e. longer survival.

```{r}
# Only uses "sex" as the predictor
fit.cox = coxph(Surv(time, status)~sex)
summary(fit.cox)  # p=0.2, no evidence for differenve in survival times by sex

# Fitting model with more predictors
fit.all = coxph(Surv(time,status)~sex+diagnosis+loc+ki+gtv+stereo)
fit.all

# Plotting survival curves for different diagnosis type
modal.data = data.frame(diagnosis = levels(diagnosis),
                        sex = rep("Female", 4),
                        loc = rep("Supratentorial", 4),
                        ki  = rep(mean(ki),4),
                        gtv = rep(mean(gtv),4),
                        stereo = rep("SRT", 4)
                        )
survplots = survfit(fit.all, newdata=modal.data)
plot(survplots, xlab="Months", ylab="Survival Probability", col=2:5)
legend("bottomleft", levels(diagnosis), col=2:5, lty=1, cex=0.7)
```

## Publication Data

```{r}
# Kaplan-Meier curves
fit.posres = survfit(Surv(time, status)~posres, data=Publication)
fit.posres
plot(fit.posres, xlab="Months", ylab=" Prob. of not being published", col=3:4)
legend("topright", c("Neg result", "Pos result"), col=3:4, lty=1, cex=0.7)

# Cox's Proportional Hazard - only 1 predictor ('positive results')
fit.pub = coxph(Surv(time, status) ~ posres, data = Publication)
fit.pub

logrank.test = survdiff(Surv(time, status)~posres, data = Publication)
logrank.test

# Cox's Proportional Hazard - all predictors
fit.pub2 = coxph(Surv(time, status) ~ . -mech, data = Publication)
fit.pub2
```

## Call Center Data - simulated

Simulated survival data using the sim.survdata() function, which is part of the 
**coxed** library. The simulated data will represent the observed wait times 
(in seconds) for 2,000 customers who have phoned a call center. In this context, 
censoring occurs if a customer hangs up before his or her call is answered.  

The **sim.survdata()** function allows us to specify the maximum possible failure 
time, which in this case corresponds to the longest possible wait time for a 
customer - set at 1,000 seconds.  

We find that differences between centers are highly significant, as are
differences between times of day.  

```{r}
set.seed(4)
N = 2000
Operators = sample(5:15, N, replace=T)
Center = sample(c("A","B","C"), N, replace = T)
Time = sample(c("Morn.", "After.", "Even."), N, replace=T)
X = model.matrix(~Operators + Center + Time)
X = X[,-1]

true.beta = c(0.04, -0.3, 0, 0.2, -0.2)

# Baseline hazard function - with one argument, representing time
h0 = function(t) return(0.00001 * t)

library(coxed)
queue = sim.survdata(N=N, T=1000, X=X, beta=true.beta, hazard.fun=h0)
names(queue)

head(queue$data)
mean(queue$data$failed) # 90% of calls were answered

# Kaplan-Meier survival curves
par(mfrow=c(1,2))

fit.Center = survfit(Surv(y, failed)~Center, data=queue$data)
plot(fit.Center, xlab="Seconds", ylab="Prob of still being on hold", col=2:4)
legend("topright", c("Center A", "Center B", "Center C"),col=2:4, lty=1, cex=0.7)
survdiff(Surv(y, failed)~Center, data=queue$data)

fit.Time = survfit(Surv(y, failed)~Time, data=queue$data)
plot(fit.Time, xlab="Seconds", ylab="Prob of still being on hold", col=2:4)
legend("topright", c("Afternoon", "Evening", "Morning"),col=2:4, lty=1, cex=0.7)
survdiff(Surv(y, failed)~Time, data=queue$data)

# Cox's Proportional Hazard
fit.queue = coxph(Surv(y, failed)~., data = queue$data)
fit.queue
```

The coefficient estimates resulting from the Cox model are fairly consistent with 
true estimates of 0.04, -0.3, 0, 0.2, -0.2.
